---
title: "Clustering"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Preprocessing


```{r }

library(plyr)
CrimeData <- read.csv("crimedata.csv")
CrimeData1 <- CrimeData
#summary(CrimeData)
#str(CrimeData)

library(dplyr)
# Rename column Community Name to remove special character - NOT WORKING 
#CrimeData <- rename(CrimeData,c("Êcommunityname"= "communityname"))

CrimeData$communityname <- CrimeData$"Êcommunityname"
#CrimeData3 <- CrimeData3[-1,]

# As community names can be duplicate for many states, NROW = 197 which means, 197 community names 
# are duplicate for many states

NROW(CrimeData[duplicated(CrimeData[c("communityname")]),])

# Lets see if the combination of communityname and state can act as a primary key
NROW(CrimeData[duplicated(CrimeData[c("communityname","state")]),])
# YES, IT CAN.. zero duplicates


# Creating a primary key column by concatenating communityname and state as "citystate"
CrimeData$citystate <- paste(CrimeData$communityname,CrimeData$state)

# Checking again to see the duplicates
NROW(CrimeData[duplicated(CrimeData[c("citystate")]),])
# DONE!! zero rows

CrimeData2 <- CrimeData

##STAGE 2

#Next step is to check variable types as per the data - numeric, character, logical, factor




CrimeData3 = sapply(CrimeData, function(x) as.numeric(format(x,decimal.part=".")) ) 
CrimeData3 = as.data.frame(CrimeData3)

#str(CrimeData3)
CrimeData3$communityname = CrimeData$communityname
CrimeData3$state = CrimeData$state
CrimeData3$citystate = CrimeData$citystate

## STAGE 3 -  handling missing values and NA's

mean(CrimeData3$ViolentCrimesPerPop)  ## NA's are there so performing any operation will not yield anything
mean(CrimeData3$ViolentCrimesPerPop,na.rm = TRUE) ## Now mean is calculated without considering NA values (preferable way)

#is.na(CrimeData3$ViolentCrimesPerPop)
#summary(CrimeData3)

# dataframe states contains an aggregated list of all the states with the mean of violent crime w.r.t population of that state
states=aggregate(x=CrimeData3$ViolentCrimesPerPop,by=list(CrimeData3$state),mean,na.rm=T)
states=arrange(states,desc(x))
#states

#head(CrimeData3)


```



```{r pressure, echo=FALSE}

aggrCrimeData2 <- aggregate(list(Population = CrimeData3$population,
                                 BlackPercentage = CrimeData3$racepctblack,
                                 WhitePercentage = CrimeData3$racePctWhite,
                                 AsianPercentage = CrimeData3$racePctAsian,
                                 Age12to21 = CrimeData3$agePct12t21,
                                 Age12to29 = CrimeData3$agePct12t29,
                                 Age16to24 = CrimeData3$agePct16t24,
                                 AgeAbove65 = CrimeData3$agePct65up,
                                 MedianHouseHold = CrimeData3$medIncome,
                                 Poverty = CrimeData3$NumUnderPov,
                                 MiddleSchoolDropOut = CrimeData3$PctLess9thGrade,
                                 HighSchoolDropOut = CrimeData3$PctNotHSGrad,
                                 BachelorOrMore = CrimeData3$PctBSorMore,
                                 Employed = CrimeData3$PctEmploy, 
                                 Unemployed = CrimeData3$PctUnemployed,
                                 Immigirants = CrimeData3$NumImmig,
                                 ImmigirantsPercentage = CrimeData3$PctForeignBorn,
                                 Homeless = CrimeData3$NumStreet,
                                 ViolentCrimes = CrimeData3$ViolentCrimesPerPop, 
                                 NonViolentCrimes = CrimeData3$nonViolPerPop,
                                 Murders = CrimeData3$murders,
                                 Rapes = CrimeData3$rapes,
                                 Robberies = CrimeData3$robberies,
                                 Assaults = CrimeData3$assaults,
                                 Burglaries = CrimeData3$burglaries,
                                 Larcenies = CrimeData3$larcenies,
                                 AuntoTheft = CrimeData3 $autoTheft),
                  by=list(state = CrimeData3$state),mean,na.rm=T)



aggrCrimeData2 <- aggrCrimeData2[ which(aggrCrimeData2$state !='DC'), ]

library(ggplot2)

ggplot(aggrCrimeData2,
       aes(x=reorder(aggrCrimeData2$state, -(aggrCrimeData2$Population/1000)),
       y = (aggrCrimeData2$Population/1000),
       fill = aggrCrimeData2$state) ) +
       geom_bar(stat = "identity", show.legend=FALSE)  +
       ggtitle("Population in the U.S states") +  
       ylab("Population (In Thousands) ") +  
       xlab("U.S State")




```

Based on population plot, grouping states into four groups with similar population size. the four group splits are in the following population range  - 
Group 1 -> 20,000 - 29,000
Group 2 -> 30,000-39,000
Group 3 -> 40,000- 69,000
Group 4 -> 69,000- 700,000



```{r}
FirstPopulationGroup <- c('VT',
         'WY',
         'ME',
         'WV',
         'NH',
         'DE',
         'NJ',
         'ID',
         'SC',
         'SD',
         'MN',
         'KY',
         'MS',
         'AR',
         'ND',
         'UT'
         )
SecondPopulationGroup<- c(
         'RI',
         'MA',
         'CT',
         'MO',
         'WI',
          'IA',
          'PA',
          'OR',
          'GA',
          'MI',
          'OH',
          'NC',
          'OK',
          'TN',
          'WA',
         'AL'
         )
ThirdPopulationGroup  <- c(
          'IN',
          'FL',
          'NM',
          'TX',
          'VA',
          'LA',
           'CO',
           'CA',
           'MD',
           'NV',
           'IL',
           'AZ',
           'NY',
           'KS',
           'AK'
          )



```

##Preparing data for kmeans clustering

Preparing data groups for clustering. Performing clustering on all the states with different population, dividinginto many clusters will limit the results. Hence performing cluster analysis on group of data with similar population size will give better clusters with similar features and makes the analysis more sensible. 

```{r}

Group1 <- aggrCrimeData2[ which(aggrCrimeData2$state %in% FirstPopulationGroup), ]
Group2 <- aggrCrimeData2[ which(aggrCrimeData2$state %in% SecondPopulationGroup), ]
Group3 <- aggrCrimeData2[ which(aggrCrimeData2$state %in% ThirdPopulationGroup), ]


```

##K-means 

Performing k means clustering using two set of feature groups. 

1. Based on unemployment, poverty and failed education and the number of violent crimes 
2. Based on type of crime - Murders,Rapes,Robberies,Assaults,Burglaries,Larcenies,AutoTheft

##Unemployment,poverty and failed education and the number of violent crimes

```{r}

#install.packages('cluster')
#install.packages('factoextra')
library(cluster)    # clustering algorithms
library(factoextra)

# Clustering first population group 

Clustergroup1 <- Group1[,-1]
rownames(Clustergroup1) <-  Group1[,1]
Clustergroup1 <- na.omit(Clustergroup1)
Clustergroup1  <- Clustergroup1[,c('Poverty', 'MiddleSchoolDropOut',
                                  'HighSchoolDropOut','Unemployed',
                                  'ViolentCrimes')]
Clustergroup1  <- scale(Clustergroup1)
Clustergroup1  <- as.data.frame(Clustergroup1)
DistanceMatrix1 <- get_dist(Clustergroup1)
DistancePlot1  <- fviz_dist(DistanceMatrix1, gradient = list(low = "#00AFBB", 
                                                            mid = "white",
                                                            high = "blue"))
KClusterGroup1 <- kmeans(Clustergroup1, centers = 2 , nstart = 30)
str(KClusterGroup1)
ClusterPlot1 <- fviz_cluster(KClusterGroup1, data = Clustergroup1, 
                             main = "Mean population < 33k")


# Clustering second  population group 

Clustergroup2 <- Group2[,-1]
rownames(Clustergroup2) <-  Group2[,1]
Clustergroup2 <- na.omit(Clustergroup2)
Clustergroup2  <- Clustergroup2[,c('Poverty', 'MiddleSchoolDropOut',
                                  'HighSchoolDropOut','Unemployed',
                                  'ViolentCrimes')]
Clustergroup2  <- scale(Clustergroup2)
Clustergroup2  <- as.data.frame(Clustergroup2)
DistanceMatrix2 <- get_dist(Clustergroup2)
DistancePlot2  <- fviz_dist(DistanceMatrix2, gradient = list(low = "#00AFBB", 
                                                            mid = "white",
                                                            high = "blue"))
KClusterGroup2 <- kmeans(Clustergroup2, centers = 2 , nstart = 30)
str(KClusterGroup2)
ClusterPlot2 <- fviz_cluster(KClusterGroup2, data = Clustergroup2,
                             main = "Mean population range 34-49k")

# Clustering third  population group 

Clustergroup3 <- Group3[,-1]
rownames(Clustergroup3) <-  Group3[,1]
Clustergroup3 <- na.omit(Clustergroup3)
Clustergroup3  <- Clustergroup3[,c('Poverty', 'MiddleSchoolDropOut',
                                  'HighSchoolDropOut','Unemployed',
                                  'ViolentCrimes')]
Clustergroup3  <- scale(Clustergroup3)
Clustergroup3  <- as.data.frame(Clustergroup3)
DistanceMatrix3 <- get_dist(Clustergroup3)
DistancePlot3  <- fviz_dist(DistanceMatrix3, gradient = list(low = "#00AFBB", 
                                                            mid = "white",
                                                            high = "blue"))
KClusterGroup3 <- kmeans(Clustergroup3, centers = 2 , nstart = 30)
str(KClusterGroup3)
ClusterPlot3 <- fviz_cluster(KClusterGroup3, data = Clustergroup3, 
                             main = "Mean population range >49k")


# plots in a grid for different groups 

library(gridExtra)
grid.arrange(DistancePlot1, DistancePlot2, DistancePlot3, 
             nrow =2, top = 'Distance matrix of the grouped population data')

grid.arrange(ClusterPlot1, ClusterPlot2, ClusterPlot3, 
             nrow = 2) 

```




##Evaluation using elbow method

Using elbow method to find the optimal number of clusters 

```{r}



set.seed(1000)
Plot1 <- fviz_nbclust(Clustergroup3,kmeans, method = "wss")

set.seed(1000)
Plot2 <- fviz_nbclust(Clustergroup2,kmeans, method = "wss")

set.seed(1000)
Plot3 <- fviz_nbclust(Clustergroup1,kmeans, method = "wss")

grid.arrange(Plot1, Plot2, Plot3, nrow = 2, top = 'Optimal number of clusters for first clustering group')


```

The elbow drops at 5 for all the groups divided hence 5 is the optimal number of clusters for splitting the data. 

Updating k = 5 in the cluster size

```{r}


Clustergroup1 <- Group1[,-1]
rownames(Clustergroup1) <-  Group1[,1]
Clustergroup1 <- na.omit(Clustergroup1)
Clustergroup1  <- Clustergroup1[,c('Poverty', 'MiddleSchoolDropOut',
                                  'HighSchoolDropOut','Unemployed',
                                  'ViolentCrimes')]
Clustergroup1  <- scale(Clustergroup1)
Clustergroup1  <- as.data.frame(Clustergroup1)
DistanceMatrix1 <- get_dist(Clustergroup1)
DistancePlot1  <- fviz_dist(DistanceMatrix1, gradient = list(low = "#00AFBB", 
                                                            mid = "white",
                                                            high = "blue"))
KClusterGroup1 <- kmeans(Clustergroup1, centers = 5 , nstart = 30)
str(KClusterGroup1)
ClusterPlot1 <- fviz_cluster(KClusterGroup1, data = Clustergroup1, 
                             main = "Mean population < 33k")


# Clustering second  population group 

Clustergroup2 <- Group2[,-1]
rownames(Clustergroup2) <-  Group2[,1]
Clustergroup2 <- na.omit(Clustergroup2)
Clustergroup2  <- Clustergroup2[,c('Poverty', 'MiddleSchoolDropOut',
                                  'HighSchoolDropOut','Unemployed',
                                  'ViolentCrimes')]
Clustergroup2  <- scale(Clustergroup2)
Clustergroup2  <- as.data.frame(Clustergroup2)
DistanceMatrix2 <- get_dist(Clustergroup2)
DistancePlot2  <- fviz_dist(DistanceMatrix2, gradient = list(low = "#00AFBB", 
                                                            mid = "white",
                                                            high = "blue"))
KClusterGroup2 <- kmeans(Clustergroup2, centers = 5 , nstart = 30)
str(KClusterGroup2)
ClusterPlot2 <- fviz_cluster(KClusterGroup2, data = Clustergroup2,
                             main = "Mean population range 34-49k")

# Clustering third  population group 

Clustergroup3 <- Group3[,-1]
rownames(Clustergroup3) <-  Group3[,1]
Clustergroup3 <- na.omit(Clustergroup3)
Clustergroup3  <- Clustergroup3[,c('Poverty', 'MiddleSchoolDropOut',
                                  'HighSchoolDropOut','Unemployed',
                                  'ViolentCrimes')]
Clustergroup3  <- scale(Clustergroup3)
Clustergroup3  <- as.data.frame(Clustergroup3)
DistanceMatrix3 <- get_dist(Clustergroup3)
DistancePlot3  <- fviz_dist(DistanceMatrix3, gradient = list(low = "#00AFBB", 
                                                            mid = "white",
                                                            high = "blue"))
KClusterGroup3 <- kmeans(Clustergroup3, centers = 5 , nstart = 30)
str(KClusterGroup3)
ClusterPlot3 <- fviz_cluster(KClusterGroup3, data = Clustergroup3, 
                             main = "Mean population range >49k")


# plots in a grid for different groups 

library(gridExtra)
grid.arrange(ClusterPlot1, ClusterPlot2, ClusterPlot3, 
             nrow = 2) 



```
